Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-09-19 review_article.Rmd
interacting and animating switch

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, fig.height = 5)
library(ggplot2)
library(grid)
library(png)
library(gridExtra)
```
<!--

@@

```{r choro, fig.cap = "Choropleth map of the average age-adjusted rate per 100,000 people, of new cases of lung and bronchus cancer 2012-2016 in the USA. Utah has an unusually low rate and Kentucky is high. There is generally a west to east trend, with higher rates towards the easern states.", message=FALSE, warning=FALSE, out.width='100%', fig.align='center', fig.width=12}
library(sugarbag)
library(tidyverse)
library(spData)
library(ggthemes)
library(maptools)
library(sf)
library(cartogram)
#knitr::include_graphics("figures/ggchoro.png")
# read in data
cancer <- read_csv("data/USCSlung.csv") %>% rename(NAME = Area)

@@

ggdraw() +
  draw_plot(rasterGrob(plot1), 0, .75, 0.33, .25) +
  draw_plot(rasterGrob(plot2), 0.33, 0.75, 0.33, .25) +
  draw_plot(rasterGrob(plot3), 0.66, 0.5, .33, .5) +
  draw_plot(rasterGrob(plot4), 0, 0.5, .33, .25) +
  draw_plot(rasterGrob(plot5), 0.33, 0.5, .33, .25) +
  draw_plot(rasterGrob(plot6), 0, 0, .66, .5) +
  draw_plot(rasterGrob(plot7), .66, 0, .33, .5) +
  draw_plot_label(c("a", "b", "c", "d", "e", "f", "g"), 
    c(0.015, 0.33, 0.63, 0.015, 0.33, 0.014, 0.63), 
    c(0.98, 0.98, 0.98, 0.74, 0.74, 0.47, 0.47), size = 15)

@@

Figure 2.a contains an image from the *[The Environmental and Health Atlas of England and Wales](http://www.envhealthatlas.co.uk/eha/Breast/)*. This map shows the relative risk for women developing lung cancer in England and Wales in 2010. The cancer data used to generate this map came from Office for National Statistics (ONS) (England) and from the Welsh Cancer Intelligence and Surveillance Unit (WCISU).

@@

The map seen in Figure 2.b is from the *[Globocan 2012: Estimated  Cancer Incidence, Mortality and Prevalence](http://globocan.iarc.fr/Pages/Map.aspx)*. This global map shows age standardized incidence rates (per 100,000) for all invasive cancers for both men and women, aggregated at a national level for 2018. This map is published by the [World Health Organisation's International Agency for Research on Cancer](https://www.iarc.fr/). Data was sourced from cancer registries in each country, contributing registries can be seen in the supplementary material on the cancer atlas website.

@@

Figure 2.c shows an extract from the *[The Atlas of Cancer in QLD](https://cancerqld.org.au/research/queensland-cancer-statistics/queensland-cancer-atlas/)*. This map was published by the Queensland Cancer Council and shows the relative incidence ratio of lung cancer in males in the state of QLD within Australia based on data from 1998 to 2007. Data to generate this map was sourced from the *[Queensland Cancer Registry](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/)*.

### Bowel Cancer Australia Atlas 
Figure 2.d shows the *[Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/)*. Published by *[Bowel Cancer Australia](https://www.bowelcanceraustralia.org/)* (Australia). This map shows the percentage of Australian males between 50 - 54 years of age that diagnosed with bowel cancer in 2016 in Australia. The source of the data is not provided.

### United States Cancer Statistics: An Interactive Cancer Statistics Website
The *[United States Cancer Statistics: An Interactive Cancer Statistics Website](https://nccd.cdc.gov/DCPC_INCA/)* can be seen in Figure 2.e. This map contains  the incidence rate per 100,000, of all cancer types for men and women in the United States in 2016, aggregated at the state level. The map was published by the *[Centers for Disease Control and Prevention](https://www.cdc.gov/)*. Incidence data seen in this map were compiled from cancer registries meeting U.S. Cancer Statistics data quality criteria covering 100% of the U.S. population.

@@

### Map of Cancer Mortality Rates in Spain
The *[Map of Cancer Mortality Rates in Spain](http://elpais.com/elpais/2014/10/06/media/1412612722_141933.html)* can be seen in Figure 2.f. The side by side maps show relative risk of lung cancer for men vs women based on data from 2004 to 2008. The source of the data and statistical methods are unknown.

### Atlas of Childhood Cancer in Ontario 
Figure 2.g shows an extract from the Atlas of Childhood Cancer in Ontario and specifically displays the *[incidence rate of childhood cancers](https://www.pogo.ca/wp-content/uploads/2015/02/POGO_CC-Atlas-3-Incidence_Feb-2015.pdf)* per 100,000 (by census division) for children aged 0-14, in Ontario from 1995 to 2004.

@@

## Interactivety

One of the concerns of  uncertainty communication to the non-expert is fear of cognitive overload [@mcgranaghan1993cartographic] in which the the user reaches an information threshold, beyond which will not be able to make sense of the information. These concerns are not unfounded, complexity and density of representation methods appear to overwhelm novice decision makers, while experts are able to use the detail more readily when making-decisions [@cliburn2002design]. Interactivety is a design feature within modern mapping methods that can be used to incorporate additional information and complexity without overloading the user.

Interactivety enables supplementary information to be incorporated into online atlases without cluttering the display. Interactive design features found in online cancer maps include tool tip features, drop down menus, data selection, zooming and panning allow users to explore the map as they want more information and allow flexibility in the display [@HTLWM]. The use of these supports were found in various online cancer maps identified by @roberts2019communication. The controls for basic interactive features are often placed outside of the plot space [@TGA], thus the map image is updated/replaced as the user interacts with the controls. For example, changing the population age or other demographic variables. Some more advanced interactions include direct interactions with the plot via the use of overlaid tool tip features, very few cancer atlases involve these more complex selection interactivities.

(insert figure of both types of interactivity)

Additionally, interactivity allows the user to toggle between different variables, map views or multiple realizations of possible future scenarios [@goodchild1994introduction]. Thus providing additional mechanisms for the users comprehension as well as the uncertainty of the available information (@maceachren1992visualizing; @van1994visualization).

These interactive features provide an opportunity for users to explore the additional information available. This helps users to understand and interpret the spatial distribution presented, as well as validate, explain or explore the presented statistics and their relationships to each other and/or their underlying spatial distribution. This allows relationships between spatial areas and diseases to be explored with sophistication in non-traditional but still 'cognitively accesible' ways [@TNTEA]. The interactive features of the publically available maps identified by @roberts2019communication allow the exploration of geographic hierarchies, population distribution, statistical uncertainty, demographics and socio-economic indicators.
@TNTEA suggested LM plots as a solution to linking cartography and statistical graphics.

(insert figures showing different interactivity)

<!--

Technological advances may provide a solution to this through the use of interactivity, animation, and sound as additional channels for representing uncertainty information without interfering with the ability to see any features that are present in the display [@maceachren1992visualizing]. However, if uncertainty information is perceived as integral and inseparable from the data it is associated with, assumptions of uncertainty creating cognitive overload may not be as valid.

@@

ggdraw() +
  draw_plot(rasterGrob(i1), 0,  0.13,   .28,  .83) + #a
  draw_plot(rasterGrob(i2), 0.32, 0.56, .68, .4) + #b
  draw_plot(rasterGrob(i3), 0.32, 0.41, .68, .13) + #c
  draw_plot(rasterGrob(i4), 0.32, 0.14, .68, .25) + #d
  draw_plot(rasterGrob(i5), 0, 0, 1, .13) + #e
  draw_plot_label(c("i", "ii", "iii", "iv", "v"), 
    c(0.01, 0.3, 0.3, 0.3, 0.01), 
    c(0.99, 0.99, 0.57, 0.40, 0.15), size = 15)

@@

```{r ggcartograms, fig.cap = "Alternative maps of the United States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each stated shape has been distorted according to the population of the state in 2015. The state of California has become much larger due to it's large population density. This draws attention to the densely populated North East region, and detracts from the less populated Mid West.", message=FALSE, warning=FALSE, out.width='100%'}
g1 <- png::readPNG("figures/ggcont.png")
g2 <- png::readPNG("figures/ggncont.png")
g3 <- png::readPNG("figures/ggdorl.png")
g4 <- png::readPNG("figures/gghexmap.png")
#knitr::include_graphics("figures/usa_grid.png")

library(cowplot)
ggdraw() +
  draw_plot(rasterGrob(g1), 0, .5, 0.5, 0.5) +
  draw_plot(rasterGrob(g2), 0.5, 0.5, 0.5, 0.5) +
  draw_plot(rasterGrob(g3), 0.0, 0.0, 0.5, 0.5) +
  draw_plot(rasterGrob(g4), 0.5, 0, 0.5, 0.5)+
  draw_plot_label(c("a", "b", "c", "d"), 
    c(0, .5, 0, 0.5), 
    c(1, 1, .5, .5), size = 15)
```

Table: \label{tab:usa} Maps used to present statistics for the Uited States of America

@@

@@

# Animation and Interactivity {#ch:interacting}

Epidemiologists consider the interactions between the spatial distribution of a disease, the environment and the people affected.
Recent developments in technology have created an opportunity to migrate the presentation of epidological results from atlases in paper map collections to web atlases.
When the map is supplementary to the discussion, static visualisations can help to aid the written results provided by experts, and pairing an explanation with a map or diagram can help to draw attention to a point, group or area of interest.
Static maps often need to include enough information for the display to be understandable out of the original context.
There are now many suitable methods and mediums to communicate health data, the choice between them should always be driven by the intended message to communicate and the needs of the audience.

@CPISACA provide three questions that are generally when using a disease map:

@@

A single map display historically created by epidemiologists, may not effectively answer all three at the same time.
This is especially true when areas are small and difficult to identify in a display.
However, with the ability to animate, map creators can specify and answer these questions by directing the attention of a map user to the areas of interest.

@@

Animations can be implemented to allow users varying degrees of control. 
A basic form of interactivty with animations is the play, pause and rewind of films, where the film is intended to be watched in one direction but the position in the animation can be controlled.
@CPISACA provide weather maps as well developed and appropriate examples of spatial displays to communicate to the general public, and encourage those creating disease maps to leverage the success of weather displays.
The movement of a weather system will follow a forecasted path, the animation will communicate this so that all map users can follow the path of the weather system across the geography over a specified period of time.
More complex animations may include steps, when user interaction via clicks animates the display between stages. 
The Australian Cancer Atlas [-@TACA] provides [tours](
https://atlas.cancer.org.au/app/tour/lungcancer) which change the display to draw user's attention to areas on the map that are relevant to the story.
This directs user's attention and gives them tools to plan their own exploration.

<!-- Interactivity (Compare movies or tv to video games) -->
The move from static, print media displays toward interactivity allow users to develop maps to ask and answer their own questions.
Where the needs of the audience is changeable and is the priority, the map creator can allow interactivity for map users to explore.

Developmets and adoptions of this technology has allowed map users more control of the message displayed.
Users are able to see the familiar shapes, zoom in to explore densely populated communities, or change the display completely to highlight the population space.

This oppportunity provides room to move away from the traditional approach of a single map for a single use, into a more explorative territory.
Interactivity allows users to drive their own exploration, controlling a display through variable controls, links, tool tips, and data selection tools [@TGA].

Where the message is most important, static or animated graphics allow control over the display and interpretation.
However before publishing, it is important for map creators to develop and then test the message being communicated, this task requires expert knowledge.

@@

@@

<!--
# Software
Commonly available tools to implement.
-->
@tidyerse for data analysis, @spData and @sf to implement plotting of spatial data.
@sugarbag to create hexagon tessellation. @ggthemes to enhance plot displays.
@cartogram for contiguous and non-contiguous cartogram displays.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.height = 5)
library(ggplot2)
library(png)
library(grid)
```
<!--

@@

```{r choro, fig.cap = "Choropleth map of the average age-adjusted rate per 100,000 people, of new cases of lung and bronchus cancer 2012-2016 in the USA. Utah has an unusually low rate and Kentucky is high. There is generally a west to east trend, with higher rates towards the easern states.", message=FALSE, warning=FALSE, out.width='100%', fig.align='center', fig.width=12}
library(sugarbag)
library(tidyverse)
library(spData)
library(ggthemes)
library(maptools)
library(sf)
library(cartogram)
# read in data
cancer <- read_csv("data/USCSlung.csv") %>% rename(NAME = Area)

@@

ggdraw() +
  draw_plot(rasterGrob(plot1), 0, .75, 0.33, .25) +
  draw_plot(rasterGrob(plot2), 0.33, 0.75, 0.33, .25) +
  draw_plot(rasterGrob(plot3), 0.66, 0.5, .33, .5) +
  draw_plot(rasterGrob(plot4), 0, 0.5, .33, .25) +
  draw_plot(rasterGrob(plot5), 0.33, 0.5, .33, .25) +
  draw_plot(rasterGrob(plot6), 0, 0, .66, .5) +
  draw_plot(rasterGrob(plot7), .66, 0, .33, .5) +
  draw_plot_label(c("a", "b", "c", "d", "e", "f", "g"), 
    c(0.015, 0.33, 0.63, 0.015, 0.33, 0.014, 0.63), 
    c(0.98, 0.98, 0.98, 0.74, 0.74, 0.47, 0.47), size = 15) +
  theme_set(theme_cowplot())

@@

Figure 2.a contains [The Environmental and Health Atlas of England and Wales](http://www.envhealthatlas.co.uk/eha/Breast/). This map shows the relative risk for women developing lunch cancer across England and Wales in 2010. The cancer data used to generate this map came from Office for National Statistics (ONS) (England) and from the Welsh Cancer Intelligence and Surveillance Unit (WCISU).

@@

The map seen in Figure 2.b is from the [Globocan 2012: Estimated  Cancer Incidence, Mortality and Prevalence](http://globocan.iarc.fr/Pages/Map.aspx). This global map shows age standardized incidence rates (per 100,000) of all cancers for both men and women at national level, for 2018. This map is published by the [World Health Organisation's International Agency for Research on Cancer](https://www.iarc.fr/). Data was sourced from cancer registries in each country, contributing registries can be seen in the supplementary material on the cancer atlas website.

@@

Figure 2.c shows the [The Atlas of Cancer in QLD](https://cancerqld.org.au/research/queensland-cancer-statistics/queensland-cancer-atlas/). This map was published by the Queensland Cancer Council and shows the relative incidence ratio of lung cancer in males in the state of QLD within Australia (1998 - 2007). Data to generate this map was sourced from the [Queensland Cancer Registry](https://cancerqld.org.au/research/cancer-registries/queensland-cancer-registry/).

### Bowel Cancer Australia Atlas 
Figure 2.d shows the [Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/). Published by [Bowel Cancer Australia](https://www.bowelcanceraustralia.org/) (Australia). This map shows the percentage of males between 50 - 54 years of age that diagnosed with bowel cancer in 2016 in Australia. The source of the data is not provided.

### United States Cancer Statistics: An Interactive Cancer Statistics Website
The [United States Cancer Statistics: An Interactive Cancer Statistics Website](https://nccd.cdc.gov/DCPC_INCA/) can be seen in Figure 2.e. This map contains  the Incidence rate per 100,000, of all cancer types for man and women in the United States for 2016, aggregated at the state level. The map was published by the [Centers for Disease Control and Prevention](https://www.cdc.gov/). Incidence data seen in this map were compiled from cancer registries meeting U.S. Cancer Statistics data quality criteria covering 100% of the U.S. population.

@@

### Map of Cancer Mortality Rates in Spain
The [Map of Cancer Mortality Rates in Spain](http://elpais.com/elpais/2014/10/06/media/1412612722_141933.html) can be seen in Figure 2.f. The side by side maps show relative risk of lung cancer for men vs women based on data from 2004 - 2008. The source of the data and statistical methods are unknown.

### Atlas of Childhood Cancer in Ontario 
Figure 2.g shows the [incidence rate of childhood cancers](https://www.pogo.ca/wp-content/uploads/2015/02/POGO_CC-Atlas-3-Incidence_Feb-2015.pdf) per 100,000, by census division, age 0-14, in Ontario, from 1995 to 2004.

@@

Modern mapping methods have the opportunity to incorporate interactive features.
This enables supplementary information to be incorporated into online atlases without cluttering the screen.
Tool tip features and drop down menus allow for user exploration with zooming and panning features to allow flexibility in the display [@HTLWM].
The use of these supports were found in a many online disease map sources.
This directly enables epidemiologists to explore spatial distributions of cancer statistics.
Basic interactivity is often implemented by controls outside of the plot space [@TGA],
this exchanges an image or set of data for another.
Some more advanced interactions include direct interactions with the plot via the use of overlaid tool tip features, very few cancer atlases involve complex selection interactivity.

These interactive features provide an opportunity for users to explore the additional information available. This helps users to understand and interpret the spatial distribution presented, as well as validate, explain or explore the presented statistics and their relationships to each other and/or their underlying spatial distribution. 
This allows relationships between spatial areas and diseases to be explored with sophistication in nontraditional but still 'cognitively accesible' ways [@TNTEA].
The interactive features of the publically available maps identified by @roberts2019communication allow the exploration of geographic hierarchies, population distribution, statistical uncertainty, demographics and socio-economic indicators.
@TNTEA suggested LM plots as a solution to linking cartography and statistical graphics.

@@

ggdraw() +
  draw_plot(rasterGrob(i1), 0,  0.13,   .28,  .83) + #a
  draw_plot(rasterGrob(i2), 0.32, 0.56, .68, .4) + #b
  draw_plot(rasterGrob(i3), 0.32, 0.41, .68, .13) + #c
  draw_plot(rasterGrob(i4), 0.32, 0.14, .68, .25) + #d
  draw_plot(rasterGrob(i5), 0, 0, 1, .13) + #e
  draw_plot_label(c("i", "ii", "iii", "iv", "v"), 
    c(0.01, 0.3, 0.3, 0.3, 0.01), 
    c(0.99, 0.99, 0.57, 0.40, 0.15), size = 15) + 
  theme_set(theme_cowplot())

@@

```{r ggcartograms, fig.cap = "Alternative maps of the United States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each stated shape has been distorted according to the population of the state in 2015. The state of California has become much larger due to it's large population density. This draws attention to the densely populated North East region, and detracts from the less populated Mid West.", message=FALSE, warning=FALSE, out.width='100%'}
usa1 <- png::readPNG("figures/ggcont.png")
usa2 <- png::readPNG("figures/ggncont.png")
usa3 <- png::readPNG("figures/ggdorl.png")
usa4 <- png::readPNG("figures/gghexmap.png")

library(cowplot)
ggdraw() +
  draw_plot(rasterGrob(usa1), 0, .5, 0.5, 0.5) +
  draw_plot(rasterGrob(usa2), 0.5, 0.5, 0.5, 0.5) +
  draw_plot(rasterGrob(usa3), 0.0, 0.0, 0.5, 0.5) +
  draw_plot(rasterGrob(usa4), 0.5, 0, 0.5, 0.5)+
  draw_plot_label(c("a", "b", "c", "d"), 
    c(0, .5, 0, 0.5), 
    c(1, 1, .5, .5), size = 15) +
  theme_set(theme_cowplot())
```

Table: \label{tab:usa} Maps used to present statistics for the United States of America

@@

```{r auscartograms, fig.cap = "Alternative maps of the United States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each stated shape has been distorted according to the population of the state in 2015. The state of California has become much larger due to it's large population density. This draws attention to the densely populated North East region, and detracts from the less populated Mid West.", message=FALSE, warning=FALSE, out.width='100%'}
aus1 <- png::readPNG("figures/aus_ggcont.png")
aus2 <- png::readPNG("figures/aus_ggncont.png")
aus3 <- png::readPNG("figures/aus_ggdorl.png")
aus4 <- png::readPNG("figures/aus_gghexmap.png")

library(cowplot)
ggdraw() +
  draw_plot(rasterGrob(aus1), 0, .5, 0.5, 0.5) +
  draw_plot(rasterGrob(aus2), 0.5, 0.5, 0.5, 0.5) +
  draw_plot(rasterGrob(aus3), 0.0, 0.0, 0.5, 0.5) +
  draw_plot(rasterGrob(aus4), 0.5, 0, 0.5, 0.5)+
  draw_plot_label(c("a", "b", "c", "d"), 
    c(0, .5, 0, 0.5), 
    c(1, 1, .5, .5), size = 15) +
  theme_set(theme_cowplot())
```

@@

# Interactivity and Animation  {#ch:interacting}
<!--
Recent developments in technology have created an opportunity to migrate the presentation of epidological results from atlases in paper map collections to web atlases.
Pairing an explanation with a map or diagram can help to draw attention to a point, group or area of interest.
Maps may form supplementary material to the discussion, in this case static visualisations can help to aid written results. 
Static maps often need to include enough information for the display to be understandable out of the original context.
-->

There are now many suitable methods and mediums to communicate health data.
The choice between them should always be driven by the intended message to be communicated and the needs of the audience.

@CPISACA provide three questions that are generally asked when using a disease map:

@@

A single map display may not effectively answer all three at the same time.
This is especially true when areas are small and difficult to identify in a display.
However, with the ability to interact with a map, or animate the display, map creators can specify and answer these questions by directing the attention of a map user to the areas of interest.

<!-- Interactivity (Compare movies or tv to video games) -->
The progress from static print media displays to interactive methods has allowed users to develop maps to ask and answer their own questions.
Where the needs of the audience is changeable and is the priority, the map creator can allow interactivity for map users to explore. 
This oppportunity provides an alternative to the traditional approach of a single map for a single use, into a more explorative territory.
Interactivity allows users to drive their own exploration, controlling a display through variable controls, links, tool tips, and data selection tools [@TGA].

Developmets and adoptions of this technology has allowed map users more control of the message displayed. Figure \ref{fig:supp-grid-create} presents the control menus for several online cancer atlases.
Users can select combinations of cancer type, gender, and other factors to answers questions relevant to them and their work.
Users can use these controls to change the layer of geography shown, such as continents or countries, states or counties. Thess display can show familiar shapes for users to orient themselves on the map, then zoom in to explore densely populated communities.

@@

Animations can be implemented to allow users varying degrees of control. 
A basic form of interactivty with animations is the play, pause and rewind of films, where the film is intended to be watched in one direction but the position in the animation can be controlled.
@CPISACA provide weather maps as well developed and appropriate examples of spatial displays to communicate to the general public, and encourage those creating disease maps to leverage the success of weather displays.
The movement of a weather system will follow a forecasted path, the animation will communicate this so that all map users can follow the path of the weather system across the geography over a specified period of time.
More complex animations may include steps, when user interaction via clicks animates the display between stages.
This approach is implemented by Google Earth 
Where the message is most important, static or animated graphics allow control over the display and interpretation.
The Australian Cancer Atlas [-@TACA] provides [tours](
https://atlas.cancer.org.au/app/tour/lungcancer) which change the display to draw user's attention to areas on the map that are relevant to the story.
This directs user's attention and gives them tools to plan their own exploration.

@@

Several R [@R] packages were used to produce this paper.
For data analysis the tidyverse package [@tidyverse] provided many useful functions.
ggplot2 package [@ggplot2] was used to create maps, with RColorBrewer package [@RColorBrewer] providing additional colour palettes and ggthemes [@ggthemes] package providing themes.
The png package [@png] was used to access png images taken from online web atlases, cowplot package [@cowplot] was used to arrange these plots into grouped images.

The following packages were used to create tranformations from the sf [@sf] geographical shapes of the states of America [@spData] and the XXX Australian Statistical Areas (Level 3):
The cartogram [@cartogram] package creates contiguous, non-contiguous and Dorling cartgorams.
Hexagon tilegram displays were created using the sugarbag package [@sugarbag].

@@
